Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-09-19 Response_to_Reviewers.tex

\documentclass[11pt,a4paper,sans]{moderncv}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')

\usepackage[document]{ragged2e}
% pour justifier

% moderncv themes
\moderncvstyle{banking}                            % style options are 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{red}                                % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
\renewcommand{\familydefault}{\rmdefault} 
%\nopagenumbers{} 
\usepackage[utf8]{inputenc} 
\usepackage[scale=0.9]{geometry}

\def \draft {1}
\usepackage{xparse}
\usepackage{ifthen}
\DeclareDocumentCommand{\comment}{o m o o o o}
{\ifthenelse{\draft=1}{
  \IfValueT{#1}{
      \textcolor{red}{\textbf{C (#1) : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
    \IfNoValueT{#1}{
      \textcolor{red}{\textbf{C : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
 }{}
}

@@

\medskip

Concerning specifically the comments made by the first referee:

\begin{itemize}
	% 1. I think it would be better if the author would focus on a single model. This way they could describe the model and the analyses in more detail. As it stands, the models are treated as black boxes. 
	\item The suggestion to focus on a single model is a qualified comment. However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies rather than with a single model. In that perspective, the ``black-box'' is seen as an advantage as it reinforces genericity. We acknowledge the fact that this choice led us to shorten the model presentations. Nevertheless, we have taken notice of your remark by improving the description of the models and their results through the addition of pseudo-codes in Appendix, result descriptions in the text and by recalling more clearly the link to the model source we used. Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.
    %We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.

   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
   \item Concerning basic illustration of the influence of initial patterns, we think that it is not possible with single runs to show a significant effect other than the direct effect of space on the localization of agents for example. It is indeed the point of our method that such tests must be done in a systematic way rather than ``eyeball validations''.
   %\comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..}[(CC)peut-être en prenant des cas extrêmes sur Sugarscape?][(JR) je regarde... $\rightarrow$ bof, on voit rien a part que les agents se repartissent differement mais qui est une consequence directe d'ajouter l'espace, en en comparant les valeurs finales des ginis, ce qu'on fait plus bas plus rigoureusement. je suggère poliment que sa suggestion est idiote.]

	% 2. It is very important to describe the models. This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern. In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project. This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
	 %\comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants? Comment ça? Bien sûr qu'il le fait (sous contrainte du taux de vacance). Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}[(JR) ah oui cest selon la vacance ; mais il y a des regimes ou ca converge jamais sous forme de cluster et ca reste un etat dynamique sans structure ; oui par contre on a du reprendre la random reloc, c'est ce que font les physiciens aussi]

@@

    \item More generally concerning the choice of these two models, we aim at helping the readers to distance themselves from the case studies and focus on the genericity of the approach. This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs. This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.

	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
	\item Furthermore concerning the choice of the Sugarscape model, we think it is a good example of a stylized model of a spatialized socio-economic system, and according to (Batty, 2005) a good metaphor of an urban system. Our choice was accordingly clarified in the paper.
%: "In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants." % pas besoin de mettre les modifs ici

	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
	%In order to clarify the concept of ``meta-parameter'' that we prefer to keep, we now propose a short definition after the first use of the term (section Methods): "In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process".
	\item We agree that the term of meta-parameters may have been confusing, and following the suggestion we replaced it with ``generator parameters''.
     \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
    \comment[CC]{d'acc.}
    \comment[MLT]{Du coup, dans nos échanges par mails on disait qu'on était ok pour changer, mais finalement cette solution vous convient? Je comprends pourquoi il/elle a tiqué sur le terme, mais c'est pas non plus hors de propos vu l'application qu'on en fait... bref, je suis partagée!}[(JR) moi ca me va de virer le meta]

	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
	\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}[(JR) il me semble que les distribs sont assez centrées et que ca revient quasiment au même ; par contre Clem si tu as pris median pour les stats et moi mean pour les plots c'est pas fou ; tu as encore le code des analyses stats ? (c'est sur le git ?)]
	\comment[JR]{c'est la merde, on prend certaines fois la mediane et certaines fois la moyenne...}

@@

	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
	\item \comment[JR]{biblio a rajouter}

	% 7. Line 238: Please explain why this particular index was chosen and discuss its properties. Connect the discussion to other indices like L1, and Linf norms, etc. Would we get qualitatively different results from different indices? 
	\item \comment[JR]{on avait teste avec le Monge ca donnait a peut pres pareil ; donc les normes alpha surement encore plus. je regarde.}

	% 8. Regarding the spatial generator: the authors should explain why this generator is relevant to these two models? 
	\item This spatial generator is relevant to both model as it produces density grids at a ``metropolitan scale'' (see detailed discussion added in Appendix and the original paper from which we used the generator), which is the scale at which both model were initially intended to be. We made this aspect more explicit.

	\item In the case of Schelling, this is the scale at which most empirical segregation indices are computed. In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty....), or to a generic landscape where a resource is grown otherwise. In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results. Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most harmful one (e.g. the least realistic, although this is the case of most Schelling models).
	\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}[(c'est le Batty, 2005)]

@@

	% 9. The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix
	\item The behavior of the spatial generator was described with more details in Appendix.

	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
	\item Yes, we allow more agents to be in the same cell. The potential density is defined by the density grid generated, and its population with the same number of agents in all grids. This distribution is the one at initialisation. If the potential density is not reached, more agents can move in the cell, otherwise it is deemed full and unavailable for further moves. The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
	%\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
	%\comment[CC]{ça va comme réponse? }[(JR) parfait]
	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc)done]

	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
	\item cf infra. 
	\item density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution. Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling. We chose to use actual distribution of European cities to constraint our density generation.
%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]

@@

	% 12. Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.
    % \comment[JR]{plutot d'accord avec lui mais parce que proficient en NetLogo, peut etre que des debutants sont plus contraints ? quelquun aurait une ref là dessus ?}
    %\comment[MLT]{En fait, c'est surtout le cas des pre-built simulations proposées dans la librairie de NetLogo. J'ai un peu cherché, je trouve pas de références qui montrent explicitement la surabondance des grilles dans les modèles hors librairie...}[(JR) ok yes en effet ! cela devrait suffire comme argument ?]
	\item We agree that NetLogo does not restrict the use on what implementation to use. However, some features are much more easier as raster or the contrary, which could influence the choice of beginners. Furthermore, models in the built-in NetLogo model library mostly rely on rasters. We clarified that statement in the paper.

	% 13. Line 301: I am not sure what the authors mean by “interesting grids”
	\item The term ``interesting grids'' was a typo and corrected, ``grids'' only was meant.

	% 14. Line 671: The same comment regarding “mythical ABMs” 
	\item The term ``mythical'' line 671 was a language mistake and corrected, ``textbook'' ABM was meant.
    %\comment[MLT]{Use "textbook ABM instead?}[(JR) yes, better]
\end{itemize}

@@

Concerning the comments made by the second referee:

%This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.

%I also have some concerns:

\begin{itemize}
	% In the abstract the authors claim that initial spatial conditions are often taken for granted. I do not agree to this statement. Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
	\item We forcefully agree with the referee in that initial spatial conditions can have a large impact on the results of the model. Maybe not all papers take them for granted, but what we meant is that compared to other parameters, initial spatial conditions are much less often explored using sensitivity analysis. Actually, we know only a few papers which have explored the impact of initial spatial conditions as a systematic criterion in a sensitivity analysis of model results. We clarified this point in the introduction.
    %\comment[JR]{il s'agirait de nuancer ici}
	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]

@@

	% Page 2 line 98: “However, this step is rarely explicit and authors tend to generalize the conclusion of the model from a single, spatial case study”. I do not understand the first part of the this statement. The steps taken are often very explicit. Due to the amount of time it takes to correctly represent all spatial data, including population densities and the variation of other population variables over space, many authors indeed restrict themselves to a single case study due to the time intensity of this task. Many agent-based models are also not valid for other areas.
	\item \comment[JR]{idem la on parle pas exactement du meme type de modeles, mais son exemple prouve notre propos : le modele tourne sur une situation unique, et on teste pas l'effet de l'espace.}
    \comment[MLT]{Bon, c'est clairement quelqu'un qui fait de l'appliqué et qui veut pas entendre notre argument ;) On peut aller dans son sens et modifier notre propos en écrivant : "However, this step is rarely explicit and authors are therefore not able to generalize the conclusion of the model to other spatial entities."}

	% The authors do not differentiate between data driven empirical models versus theory-driven models. I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
	\item Concerning the remarks on synthetic population generation and the different types of models, we clearly stated our positioning regarding models built from theory, for which the proposed sensitivity analysis workflow is more suited and that our two case studies illustrate.\comment[JR]{oui du coup en disant ça ça repond à ses deux points précédents}\comment[JR]{positionement a plus souligner dans le papier}

	% In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models. This is contrary to the statement in the abstract that initial conditions are often taken for granted.
	\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled. But first, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition. Second, the references cited mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but none provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper. We have tried to argue this point at the end of the section.
    %\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]

@@

	% In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions. Actually this came as a surprise to me. I had the impression that all phase diagrams would be evaluated against all other diagrams. Perhaps this should be explained earlier in the paper.
	\item \comment[JR]{ce qu'il suggère c'est un indice intégré sur l'ensemble des phases diagramme, donc un ``meta-meta-indice'' en quelque sorte, nous on reste à un niveau de meta seulement car on regarde juste pour chaque s'il bouge par rapport à un point de reference. En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli. $\rightarrow$ a voir comment justifier ca bien}

	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.

	\item Color scale for distance was reversed.
    \comment[JR]{Clem tu es d'accord là dessus ?}[(CC) Non mais il chipote complètement à ce niveau! Vu la figure pour moi c'est totalement équivalent dans un sens et dans l'autre, le tout c'était qu'il y'a une double descente chromatique et y'a même pas de densité = rouge ou quoi puisque c'est un rapport. Si tu veux lui faire plaisir on change mais je vois pas trop la nécessité.][(JR) bon je vais refaire la figure pour caresser dans le sens du poil - TODO]

	% I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”). Personally I do not think that releasing the density grids is an important opportunity. As mentioned before I think that the core of the work is in the comparison of the phase vectors. I am not so sure the second opportunity listed in 4.2 is very valid. I do not think that the CEUS journal targets social sciences exactly.
	\item Thanks for the remarks on discussion, in particular on releasing density grids. We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
\end{itemize}

@@

	% In places the English along with the overall argument could be better articulated. For example, on page 1 you write: “We think simulation can become a very good tool to achieve this,”  Could this be rewritten or better argued. You are not the only one to believe this but you don’t show this in your paper.
	\item We have worked towards improving the English along with the overall argument. 
      \comment[CC]{ok pour cette phrase + derniere relecture a faire}

	% Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense. What it should be is “Generally, sensitivity analyses have focused either on the spatial context or the heterogeneity of space” or something like that. Basically you need the “or”.
	\item Thank you for the suggestion on page 4. We have modified the paragraph accordingly.

	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?  
	\item 
    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature... vous avez des références?]

### 2018-09-19 paper.tex
.

\begin{abstract}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create density grids with which spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change. 
\end{abstract}

\begin{keyword}
Space \sep Initial conditions \sep Sensitivity \sep ABM \sep Validation
\end{keyword}

@@

In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape). In no way do we want to provide a full exploration of these two particular models, their attractors and/or potential policy implications. Instead, we present a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation. The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations. The purpose is two-fold: (i) to test the robustness of simulation results to small variations of meta-parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

@@

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}. The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns. The sensitivity analysis relates to the model's parameters to how the density spatial distribution was generated and the pattern of density generated in order to produce qualitative knowledge from the model and its behaviour. In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space. In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] 
\begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{\textbf{General workflow.}} \label{fig:method}
\end{center}
\end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

@@

To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature. We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$. Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration \footnote{in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself}. The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given meta-parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidian distance. The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we sum these relative distances over the components.

The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file. In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}. \added{We use therefore OpenMOLE's functionalities for model embedding through workflow, design of experiments (parameter sampling) and high performance environment access.}

@@

The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}. An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily in our workflow. Grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{2017arXiv170806743R}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.

@@

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). We sample the meta-parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy. We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}. It yields a subset of 170 \deleted{interesting} grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg). This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results. We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model (cf. section \ref{sec:qualResults}).

>>>>

\begin{abstract}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic \added{spatial configuration} generator \deleted{controlled by meta-parameters} in order to create density grids with which spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change. 
\end{abstract}

\begin{keyword}
Space \sep Initial conditions \sep Sensitivity \sep ABM
\end{keyword}

@@

In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape). In no way do we want to provide a full exploration of these two particular models, their attractors and/or potential policy implications. Instead, we present a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation. The generator being controlled by \replaced{it own}{meta-} parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations. The purpose is two-fold: (i) to test the robustness of simulation results to small variations of \replaced{generator}{meta-} parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model. Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies. In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.

@@

The general method workflow of our method is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters ``generator parameters'' to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}. The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns. The sensitivity analysis relates to the model's parameters to how the density spatial distribution was generated and the pattern of density generated in order to produce qualitative knowledge from the model and its behaviour. In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space. In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] 
\begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{\textbf{General workflow.}} \label{fig:method}
\end{center}
\end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

@@

To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature. We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$. Phase diagrams are denoted by \replaced{the same}{a common} function $\mu$ indexed by \replaced{the generator parameters}{what we call \emph{meta-parameters}} $\vec{\alpha}$, \replaced{which}{that} capture the spatial configuration \added{(in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself)}. The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given \replaced{generator}{meta-} parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidian distance. The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we sum these relative distances over the components.

The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, this tool is a powerful way to embed both the sensitivity analysis and the \deleted{meta-}sensitivity analysis \added{to initial spatial conditions}, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file. In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}. \added{We use therefore OpenMOLE's functionalities for model embedding through workflow, design of experiments (parameter sampling) and high performance environment access.}

@@

The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}. An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily in our workflow. Grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches). To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside. The procedure stops when a fixed number of steps $t_f$ is reached. The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps). Grids are thus generated from the combination of the values of these four \replaced{generator}{meta-} parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment. We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$. Typical value ranges for the  parameters will be taken as, following \citet{raimbault2018calibration}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$. We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.

@@

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). We sample the \replaced{generator}{meta-} parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy. We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}. It yields a subset of 170 \deleted{interesting} grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg). This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results. We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model (cf. section \ref{sec:qualResults}).

\added{The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be. In the case of Schelling's segregation model for example, this scale is the one at which most empirical segregation indices are computed and compared to the model outputs. In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources \citep{batty2005agents}, or to a generic landscape where a resource is grown otherwise. In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density and that acknowledging this diversity might leads to a variation in the model outputs. Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most unrealistic one, although unfortunately the most common one in Schelling implementations.}
